\chapter{Area}



\section*{Surface integral}

Let $\Sigma$ be a smooth surface and $h\:\Sigma\to\RR$ be a smooth function.
Let us define the integral $\int_R h$ of the function $h$ along a region $R\subset \Sigma$.
(This definition can be applied to any Borel set of $R\subset \Sigma$.)

First assume that there is a chart $(u,v)\mapsto s(u,v)$ of $\Sigma$ defined on an open set $U\subset\RR^2$ such that $R\subset s(U)$.
In this case set
\[\int_R h\df \iint_{s^{-1}(R)} h\circ s(u,v)\cdot |\tfrac{\partial s}{\partial v}(u,v)\times\tfrac{\partial s}{\partial u}(u,v)|  \cdot du\cdot dv.\eqlbl{eq:area-def}\]
By the substitution rule for multiple variables (\ref{thm:mult-substitution}), the right hand side in  \ref{eq:area-def} does not depend on choice of $s$.
That is, if $s_1\:U_1\to \Sigma$ is another chart such that $s_1(U_1)\supset R$, then 
\[\iint_{s^{-1}(R)} h\circ s\cdot |\tfrac{\partial s}{\partial v}\times\tfrac{\partial s}{\partial u}|  \cdot du\cdot dv=\iint_{s_1^{-1}(R)} h\circ s_1\cdot |\tfrac{\partial s_1}{\partial v}\times\tfrac{\partial s_1}{\partial u}|  \cdot du\cdot dv.\]
In other words, definition \ref{eq:area-def} makes sense.

A general region $R$ can be subdivided into regions $R_1,R_2\dots$ such that each $R_i$ lies in the image of some chart.
After that one could define the integral along $R$ as the sum
\[\int_Rh=\int_{R_1}h+\int_{R_2}h+\dots\]


\section*{Area}

The area of a region $R$ in a smooth surface $\Sigma$ is defined as the surface integral 
\[\area R=\int_R 1.\]

\begin{thm}{Exercise}
Prove the following identity 
\[|\vec a\times \vec b|^2=\langle \vec a,\vec a\rangle\cdot\langle \vec b,\vec b\rangle-\langle \vec a,\vec b\rangle^2\]
for any vectors $\vec a,\vec b\in\RR^3$.

Use this identity to show that if the region $R$ of the surface $\Sigma$ is covered by a chart $s\:U\to \Sigma$, then
\[\area R=\iint_{s^{-1}(R)} \sqrt{\langle\tfrac{\partial s}{\partial u},\tfrac{\partial s}{\partial u}\rangle\cdot\langle\tfrac{\partial s}{\partial v},\tfrac{\partial s}{\partial v}\rangle -\langle\tfrac{\partial s}{\partial u},\tfrac{\partial s}{\partial v}\rangle^2}\cdot du\cdot dv.\]
\end{thm}

\parbf{Remark.}
The notion of area of surface is closely related to length of curve.
However, to define length we uses a different idea --- it was defined as the least upper bound on the lengths of inscribed polygonal lines.
It turns out that analogous definition does not work even for very simple surfaces;
section \ref{sec:schwarz-boot} describes the so called \emph{Schwarz's boot} --- a classical example of that type.

\section*{Flux and area}

Let $\vec u$ be a vector field defined on a smooth oriented surface $\Sigma$ with unit normal filed $\Norm$.
Recall that \emph{flux} of $\vec u$ thru $\Sigma$ is defined by the integral 
\[\flux_{\vec u}\Sigma=\int_\Sigma \langle \vec u,\Norm\rangle.\]

\begin{thm}{Observation}\label{obs:flux}
Let $\vec u$ be a vector field defined on a smooth oriented surface $\Sigma$.
Assume $|\vec u|\le 1$ at every point.
Then 
\[\flux_{\vec u}\Sigma \le \area\Sigma.\]
\end{thm}

\parit{Proof.}
Since $|\vec u|\le 1$ and $|\Norm|=1$ at every point,
we have that
\[\langle\vec u,\Norm\rangle\le 1\]
Therefore 
\[\flux_{\vec u}\Sigma=\int_\Sigma \langle\vec u,\Norm\rangle\le \int_\Sigma1=\area\Sigma.\]
\qedsf

\begin{thm}{Exercise}
Let $\Sigma$ be a compact smooth surface with boundary $\partial \Sigma$ in the $(x,y)$-plane.
Denote by $\Delta$ the compact region in the $(x,y)$-plane bounded by the $\partial \Sigma$.
Suppose that $\vec i,\vec j,\vec k$ is the standard basis in the the space.

\begin{enumerate}[(a)]
\item\label{ex:divergence}
Assume that all the points of $\Sigma$ lie in the upper half-space of the $(x,y)$-plane.
Use the divergence theorem (\ref{thm:div}) and the observation for the constant vector field $\vec k$
to show that 
\[\area\Sigma\ge\area\Delta.\]

\item\label{ex:curl} 
Use the curl theorem (\ref{thm:curl}) to prove the inequality in \ref{ex:divergence} for arbitrary smooth surface $\Sigma$ with boundary $\partial \Sigma$ in the $(x,y)$-plane, without assuming that the remaining points lie in the upper half-space.
\end{enumerate}

\end{thm}

Further we denote by $\vec i,\vec j,\vec k$ the standard basis in a $(x,y,z)$-coordinate system of $\RR^3$.

\begin{thm}{Exercise}
Let $\Sigma$ be a smooth closed surface that lies between parallel planes on distance $1$ from the $(x,y)$-plane.
Suppose that $\Sigma$ bounds a region $R$.
Use the divergence theorem and the observation for the vector field $\vec u=z\cdot \vec k$ to  show that 
\[\area \Sigma>\vol R.\]
\end{thm}

\begin{thm}{Exercise}
Let $\Sigma$ be a smooth closed surface that lies inside the infinite cylinder $x^2+y^2\le 1$.
Suppose that $\Sigma$ bounds a region $R$.
Use the divergence theorem and the observation for the vector field $\vec i,\vec j,\vec k$ to show that  
\[\area \Sigma>2\cdot\vol R.\]
\end{thm}

\begin{thm}{Corollary}
Let $\Sigma$ and $\Sigma'$ be a compact surfaces  with identical boundary lines;
that is, $\partial \Sigma=\partial \Sigma'$.
Suppose that $\Sigma$ is oriented and its unit normal field $\Norm$
can be extended to a vector field $\vec u$ such that $|\vec u|= 1$ at every point
and $\div \vec u= 0$ at every point between $\Sigma$ and $\Sigma'$. %???
Then 
\[\area\Sigma'\ge \area\Sigma.\]
\end{thm} %???is it needed???

The vector field $\vec u$ as in the corollary is called \emph{calibration} of $\Sigma$

\section*{Mean curvature and divergence}

The following lemma will be used to construct unit vector fields with vanishing or positive divergence.

\begin{thm}{Lemma}\label{lem:div+H}
Let $\Sigma$ be a smooth orented surface, $\Norm$ be the unit normal field on $\Sigma$.
Suppose that a unit vector field $\vec u$ is a smooth extension of $\Norm$ in a neighborhood of $\Sigma$.
Then at any point $p\in\Sigma$,
\[\div_p \vec u+H(p)=0,\]
where $H(p)$ is the mean curvature of $\Sigma$ at $p$.
\end{thm}

\parit{Proof.}
Consider the tangent-normal coordinates $(x,y,z)$ of $\Sigma$ at $p$ such that the $x$-  and $y$-coordinates point in the principle directions of $\Sigma$.
If $\vec i,\vec j,\vec k$ denotes the standard basis, then 
\begin{align*}
S_p(\vec i)&=k_1\cdot \vec i,
&
S_p(\vec j)&=k_2\cdot \vec j,
\end{align*}
where $S_p$ stands for the shape operator of $\Sigma$ at $p$.

Proposition~\ref{prop:S=-D} implies that
\[
\frac{\partial\vec u}{\partial x}=\frac{\partial\Norm}{\partial x}=-S_p(\vec i),
\quad\text{and}\quad
\frac{\partial\vec u}{\partial y}=\frac{\partial\Norm}{\partial y}=-S_p(\vec j).
\]
It follows that
\[
\langle\tfrac{\partial\vec u}{\partial x},\vec i\rangle =-k_1, 
\quad\text{and}\quad
\langle\tfrac{\partial\vec u}{\partial y},\vec j\rangle=-k_2. \eqlbl{eq:divxy}
\]


Further, since $\vec u$ is unit, we have $\langle\vec u,\vec u\rangle\equiv 1$.
Taking derivative of this identity, we get
\begin{align*}
0&=\tfrac{\partial}{\partial z} \langle\vec u,\vec u\rangle=
\\
&=2\cdot \langle\tfrac{\partial}{\partial z}\vec u,\vec u\rangle.
\end{align*}
In particular,
\[\langle\tfrac{\partial}{\partial z}\vec u,\vec k\rangle_p=0.\eqlbl{eq:divz}\]

Let us write the field $\vec u$ in the standard basis 
$\vec u=u_1\cdot \vec i+u_2\cdot\vec j+u_3\cdot\vec k$.
Since the basis $\vec i,\vec j,\vec k$ is orthonormal, the functions $u_1,u_2,u_3$ can be defined by
\begin{align*}
u_1&=\langle\vec u,\vec i\rangle,
&
u_2&=\langle\vec u,\vec j\rangle,
&
u_3&=\langle\vec u,\vec k\rangle.
\end{align*}
Applying the definition of divergence and using \ref{eq:divxy} and \ref{eq:divz}, we obtain
\begin{align*}
\div\vec u&=\tfrac{\partial u_1}{\partial x}+\tfrac{\partial u_2}{\partial y}+\tfrac{\partial u_3}{\partial z}=
\\
&=\tfrac{\partial}{\partial x}\langle\vec u,\vec i \rangle+\tfrac{\partial}{\partial y}\langle\vec u,\vec j \rangle+\tfrac{\partial}{\partial z}\langle\vec u,\vec k\rangle=
\\
&=\langle\tfrac{\partial\vec u}{\partial x},\vec i \rangle+\langle\tfrac{\partial\vec u}{\partial y},\vec j \rangle +\langle\tfrac{\partial\vec u}{\partial z},\vec k \rangle
\\
&=-k_1(p)-k_2(p)+0
\\
&=-H(p)
\end{align*}
\qedsf

Let $V$ be a body in $\RR^3$ bounded by a closed smooth surface $\Sigma$;
assume $\Sigma$ is equipped with orientation defined by unit normal field $\Norm$ that points outside $V$.
We say that $V$ is \emph{mean-convex} if the mean curvature of $\Sigma$ is nonpositive.

\begin{thm}{Exercise}\label{ex:mean-convex}
Let $V$ be a mean-convex body in $\RR^3$ bounded by a closed smooth surface $\Sigma$.
Denote by $W$ the outer region of $\Sigma$, it is a complement of the interior of $V$.

\begin{enumerate}[(a)]
\item\label{ex:mean-convex:u} 
Suppose $V$ is star-shaped; that is, there is a point $p\in V$ such that for any other point $x\in V$ the line segment $[p,x]$ lies in $V$.

Construct a unit vector field $\vec u$ on $W$ such that $\div  \vec u\ge 0$ at every point in $W$ and the restriction of $\vec u$ to $\Sigma$ is a normal field that points in $W$.
 
\item\label{ex:mean-convex:area} Suppose that another body $V'$ is bounded by a closed smooth surface $\Sigma'$ and $V'\supset V$.
Use part \ref{ex:mean-convex:u} and the divergence theorem to show that 
\[\area\Sigma'\ge \area\Sigma\]
if $V$ is star-shaped.

\item Construct a non-star-shaped mean-convex body $V$ bounded by a smooth surface such that the inequality in\ref{ex:mean-convex:u} does not hold for some body $V'\supset V$ with smooth boundary $\Sigma'$.
\end{enumerate}
\end{thm}

\section*{Area minimizing surfaces are minimal}

A smooth surface $\Sigma$ is called \emph{area minimizing} if for any compact surfaces $\Delta$ in $\Sigma$ with boundary line $\partial \Delta$ 
the following inequality 
\[\area \Delta\le \area \Delta'\]
holds for any other smooth compact surface $\Delta'$ with the same boundary line; that is, if $\partial\Delta'\z=\partial\Delta$.

Recall that a surface with vanishing mean curvature is called \emph{minimal}.


\begin{thm}{Proposition}\label{prop:minimizing-is-minimal}
Any area minimizing surface is minimal.
\end{thm}



\parit{Proof.}
Assume $\Sigma$ is a surface with nonzero mean curvature at some point $p$.
Without loss of generality we may assume that it is positive,
otherwise switch the orientation.

Let $z=f(x,y)$ be a local description of $\Sigma$ in the tangent-normal coordinates at $p$.
Denote by $H(x,y)$ and $\Norm(x,y)$ the mean curvature and the unit normal vector of the graph at the point $(x,y,f(x,y))$.
Passing to a smaller domain of $f$, we can assume that 
\[H(x,y)>0\] for any $(x,y)\in\Dom f$.

Consider the vector field $\vec u$ on the domain $\Omega=\RR\times\Dom f$ defined by 
$\vec u(x,y,z)=\Norm (x,y)$.
Note that 
\[(\div \vec u)(x,y,z)+H(x,y)=0.
\eqlbl{div<0}\]
Indeed, by construction, $\vec u$ is invariant with respect to shifts of $\Omega$ up or down.
In particular the divergence $\div \vec u$ does only on $x$ and~$y$.
Therefore it is sufficient to show \ref{div<0} for points on the graph $z=f(x,y)$.
The latter follows from Lemma~\ref{lem:div+H}.

Fix a closed $\eps$-neighborhood $D_\eps$ of the origin in the $(x,y)$-plane;
we can assume that $D_\eps$ lies in the domain of $f$.
Choose a smooth function $(x,y)\mapsto h(x,y)$ defined on $D_\eps$ in its interior and vanishes on its boundary;
for example, $h=\eps^2-x^2-y^2$ will do.
Set
\[f_t(x,y)=f(x,y)+t\cdot h(x,y).\]
Denote by $\Delta_t$ be the graph $z=f_t(x,y)$.
Set $a(t)=\area\Delta_t$ and $b(t)\z=\flux_{\vec u} \Delta_t$.
Observe that both functions $t\mapsto a(t)$ and $t\mapsto b(t)$ are smooth.

By construction of $\vec u$, we have that $a(0)=b(0)$.
By Observation \ref{obs:flux}, we have that $a(t)\ge b(t)$ for any $t$.
Therefore
\[a'(0)=b'(0).\eqlbl{eq:a'=b'}\]

Fix $t>0$.
Let $\Theta_t$ be the domain squeezed between $\Delta$ and $\Delta_t$;
that is, 
\[\Theta_t=\set{(x,y,z)}{f(x,y)< z< f_t(x,y)}.\]

By divergence theorem, we have 
\begin{align*}
b(t)-b(0)&=\iiint_{\Theta_t}\div \vec u\cdot dx\cdot dy \cdot dz=
\\
&=-\iiint_{\Theta_t} H(x,y)\cdot dx\cdot dy \cdot dz=
\\
&=t\cdot\left[- \iint_\Delta H(x,y)\cdot h(x,y) dx\cdot dy\right].
\end{align*}
Since $H>0$ and $h>0$ in the interior of $\Delta$,
the last integral is positive.
It follows that $b(t)$ is a linear function with negative slope.
By \ref{eq:a'=b'}, $a'(0)=b'(0)< 0$.
In particular, 
\[\area\Delta_t<\area \Delta\] for small $t>0$;
that is, $\Sigma$ is not area minimizing.
\qeds

\section*{Area minimizing surfaces}

The following two exercises show that minimal surface might be not area minimizing.
Recall that catenoid and helicoid are minimal; see exercises \ref{ex:catenoid-is-minimal} and \ref{ex:helicoid-is-minimal}.
The following exercise state that sufficienlty large piece of these surfaces are not area minimizing.

\begin{thm}{Exercise}
Show tha the \emph{catenoid}
\[(\cosh z)^2=x^2+y^2.\]
is not area minimizing.
\end{thm}

\begin{thm}{Exercise}
Show that the \emph{helicoid} 
\[s(u,v)=(u\cdot \sin v,u\cdot \cos v,v).\]
 is not a area minimizing.
\end{thm}

\begin{thm}{Theorem}
Let $\Sigma$ be a graph $z=f(x,y)$ of a smooth function $f$ defined on an open convex set in the $(x,y)$-plane.
Suppose $\Sigma$ is minimal, then it is area minimizing.

In particular, any minimal surface $\Sigma$ is \index{locally area minimizing surface}\emph{locally area minimizing};
that is, some neighborhood of any point $p$ in $\Sigma$ is area minimizing.
\end{thm}

The proof of the theorem is built on the following lemmas which are interesting on their own.

\begin{thm}{Lemma}
Let $D$ be a closed domain in the $(x,y)$-plane bounded by a closed piecewise smooth simple curve $\gamma$,
and $f$ a smooth function defined on $D$.
Let $\Delta$ be the graph $z=f(x,y)$ and $\Delta'$ be another piecewise smooth surface that has the same boundary as $\Delta$ and lies in the cylinder $D\times \RR$.
Then
\[\area\Delta'\ge\area \Delta.\]
Moreover in case of equality, we have $\Delta'=\Delta$.
\end{thm}

\parit{Proof.}
Note that $\Delta'$ is oriented.
Indeed, by ???, the complement of $\Delta'$ in $D\times \RR$ has two components (lower and upper), so we can equip $\Delta'$ with a unit normal vector field $\Norm'$ that points in upper component. 

Let us denote by $\Norm(x,y)$ the unit normal vector at the point $(x,y,f(x,y))\in\Delta$ that points up;
that is, its $z$-coordinate is positive at each point.
Consider the vector field $\vec u$ on the domain $\RR\times D$ defined by 
$\vec u(x,y,z)\z=\Norm (x,y)$.
By construction $|\vec u|\equiv 1$ and according to Lemma~\ref{lem:div+H}, $\div \vec u\equiv -H\equiv 0$.
Therefore $\vec u$ is a calibrating $\Delta$ in $\RR\times D$.
In particular
\[\begin{aligned}
\area \Delta&=\flux_{\vec u}\Delta,
\\
\area \Delta'&\ge\flux_{\vec u}\Delta'.
\end{aligned}
\eqlbl{area=>=flux}\]

Observe that $\RR\times D$ is contactable???.
By ??? it follows that there is a vector field $\vec v$ such that $\curl \vec v=\vec u$.
Since $\Delta$ and $\Delta'$ share the boundary, the curl theorem (???), implies that 
\[\flux_{\vec u}\Delta=\flux_{\vec u}\Delta'.\]
Finally observe that the last equality and \ref{area=>=flux} imply the main statement in the lemma.

Recall that
\[\flux_{\vec u}\Delta'\df\iint_{\Delta'}\langle\vec u,\Norm'\rangle.\]
Therefore if the equality the second line of \ref{area=>=flux}, then normal $\vec u=\Norm'$ everywhere on $\Delta'$.
In particular $\Delta'$ is a graph of 
\qeds

If the both surfaces $\Delta$ and $\Delta'$ would be oriented,
then the statement can be proved the same way as previous lemma.
However, it might be not the case.
While reading the proof it would be useful to keep the following example in mind.

\begin{thm}{Exercise}
Construct a smooth embedded M\"obius strip with unit circle as a boundary. 
\end{thm}

\begin{thm}{Lemma}
Let $f$ be a function defined on a open domain $\Omega$???
Suppose that $\Delta$ is a surface in the graph $z=f(x,y)$ bounded by a smooth closed curve $\gamma$ and $\Delta'$ be another surface whtih the same boundary.
Then
\[\area\Delta'\ge\area \Delta.\]
Moreover in case of equality, we have $\Delta'=\Delta$.
\end{thm}



\section*{Monotonicity}

\begin{thm}{Proposition}
Let $\Sigma$ be a area minimizing surface with boundary curve $\gamma$.
Suppose $p\in \Sigma$ and $R$ is the distance from $p$ to $\gamma$.
Given 
\[f\:r\mapsto \frac{\area \Sigma_r}{r^2}\]
is a nondecreasing function for $r\in (0,R]$;
here $\Sigma_r=\Sigma\cap \cap B(p,r)$.
\end{thm}

\parit{Proof.}
Without loss of generality we may assume that $p$ is the origin.

According to ???, for almost all $r$, the intersection of the $r$-sphere centered at $p$ with $\Sigma$ is a collection of smooth curves.
Denote by $\ell(r)$ the total length of this collection.

Fix small $\eps>0$.
Let us modify the surface $\Sigma$ in the ball $B(p,r)$ the following way.
Consider the rescaled copy of the intersection $\Sigma_r$ with coefficient $1-\eps$ and a conic collar 
formed by the line segments from $(1-\eps)\cdot x$ to $x$ for all points $x\in\gamma$.
Note that the obtained piecewise smooth surface $\Sigma'$ has boundary $\gamma$, the same as the boundary of $\Sigma$.

Denote by $a(r)$ the area of $\Sigma_r$.
Note the rescaled copy of $\Sigma_r$ has area $(1-\eps)^2\cdot a(r)$
and the area of collar is $(1-\tfrac\eps2)^2\cdot\ell\cdot r$.

Since $\Sigma$ is area minimizing, we have
\[\area\Sigma'\ge \area \Sigma.\]
The latter is equivalent to 
\[(1-\eps)^2\cdot a+(1-\tfrac\eps2)^2\cdot\ell\cdot r\ge a.\]
Applying this inequality for $\eps\to 0+$, we get that
\[\ell(r)\cdot r\ge 2\cdot a(r).\]

According to the coarea formula, 
\[a'(r)\ge \ell(r)\]
if the left and right hand sides are defined at $r$.
Whence 
\[a'(r)\cdot r\ge 2\cdot a(r).\]





