\chapter{Homework assignments}

\parbf{HWA-01.}
Exercises: 
\ref{ex:B2inB1}, 
\ref{ex:L-shape}, 
\ref{ex:cycloid}, 
\ref{ex:viviani}, 
\ref{ex:intrinsic-convex}.

\parbf{HWA-02.}
Exercises: 
\ref{ex:nonrectifiable-curve}(\ref{ex:nonrectifiable-curve:b}), 
\ref{ex:arc-length-helix}, 
\ref{ex:convex-hull},  
\ref{ex:S1-intrinsic},  
\ref{ex:flaw}.

\parbf{HWA-03.}
Exercises:
\ref{ex:S2-intrinsic},
\ref{ex:antipodal},
\ref{ex:bisection-of-S2}, 
\ref{ex:curvature-formulas}\ref{ex:curvature-formulas:a},    
\ref{ex:helix-curvature}.


\parbf{HWA-04.}
Exercises:
\ref{ex:curvature-of-spherical-curve},
\ref{ex:curvature-graph},
%\ref{ex:chord-lemma-optimal},
\ref{ex:sef-intersection},
\ref{ex:quadrisecant} 
+
1.12.17 in the Toponogov's book.

\parbf{HWA-05.}
Exercises:
\ref{ex:length>=2pi},
\ref{ex:chord-lemma-optimal},
\ref{ex:monotonic-tc},
%\ref{ex:schwartz},
\ref{ex:lancret},
\ref{ex:zero-tsc}.

\parbf{HWA-06.}
Exercises:
\ref{ex:schwartz},
\ref{ex:3D-spiral},
\ref{ex:double-tangent},
\ref{ex:vertex-support},
\ref{ex:between-parallels}.

\parbf{MIDTERM.} 5 problems, 20 points each:
\begin{itemize}
\item 2 theorems from the following list; in the brackets I give the corresponding statement in the Toponogov's book, if it exists.\\
\ref{lem:perimeter},
\ref{thm:length-semicont},
\ref{lem:hemisphere} (Prob. 1.10.4),\\
\ref{prop:a'-pertp-a''},
\ref{thm:fenchel} (Thm. 1.10.1),
\ref{thm:gen-fenchel},
\ref{lem:chord} (Lem. 1.11.2),
\ref{prop:inscribed-total-curvature},
\ref{prop:fenchel=},
\ref{lem:bow} (Prob. 1.10.6),
\ref{thm:fund-curves} (Thm. 1.9.1),\\
\ref{prop:total-signed-curvature} (Prob. 1.7.7),
\ref{lem:spiral},
\ref{prop:supporting-circline} (Prob. 1.7.1),
\ref{lem:lens},
\sout{\ref{thm:moon}}.
\item 2 problems from HWA's;
\item 1 new problem.
\end{itemize}

\noindent{\rule{\textwidth}{.1mm}}


\parbf{HWA-07.}
Exercises:
\ref{ex:length-dist},
\ref{ex:spherical-frenet}(\ref{ex:spherical-frenet:tau}+\ref{ex:spherical-frenet:nu}),
\ref{ex:in-triangle},
\ref{ex:line-curve-intersections}(\ref{ex:line-curve-intersections:a}+\ref{ex:line-curve-intersections:b}),
\ref{ex:finish-moon}.

\parbf{HWA-08.}
Exercises:
\ref{ex:spherical-frenet}(finish),
\ref{ex:moon-loop},
\ref{ex:inversion},
\ref{ex:revolution},
\ref{ex:hyperboloinds}.

\parbf{HWA-09.}
Exercises:
\ref{ex:tangent-chart},
\ref{ex:tangent-normal},
\ref{ex:line-of-curvature},
\ref{ex:gauss+orientable},
\ref{ex:positive-gauss}.

\parbf{HWA-10.}
Exercises:
\ref{ex:surrounds-disc},
\ref{ex:meusnier},
\ref{ex:principle-revolution},
\ref{ex:saddle-linear},
\ref{ex:one-side-bernshtein}.

\parbf{HWA-11.}
Exercises:
\ref{ex:moon-revolution}, 
\ref{ex:length-of-bry},
\ref{ex:circular-cone-saddle},
\ref{ex:convex-surf},
\ref{ex:circular-cone} 
+ try to think about this \ref{ex:lasso}.

\parbf{HWA-12.}
Exercises:
\ref{ex:small-gauss},
\ref{ex:hat-convex},
\ref{ex:intrinsic-diameter},
\ref{ex:reflection-geodesic},
\ref{ex:two-min-geod}.
%\ref{ex:min-geod+plane}.
