\chapter{Surfaces}

\section*{General definition}

Few times we will need the following general definition.

A path connected subset $\Sigma$ in a metric space is called \emph{surface} (more precisely \emph{embedded surface without boundary}) 
if any point of $p\in \Sigma$ admits a neighborhood $W$ in $\Sigma$ which is \emph{homeomorphic} to an open plane subset;
that is, if there is an injective continuous map $U\to W$ from an open set $U\subset \RR^2$ such that its invese $W\to U$ is also continuous.

However, as well as in the case of curves we will be mostly interested in smooth surfaces in the Euclidean space describe in the following section.

\section*{Smooth surfaces}

Recall that a function $f$ of two variables $x$ and $y$ is called \emph{smooth} if all its partial derivatives $\frac{\partial^{m+n}}{\partial x^m\partial y^n}f$ are defined and are continuous in the domain of definition of $f$. 

A path connected set $\Sigma \subset \mathbb{R}^3$ is called a \emph{smooth surface} (or more precisely \emph{smooth regular embedded surface}) if it can be described locally as a graph of a smooth function in an appropriate coordinate system.\label{page:def-smooth-surface}

More precisely, for any point $p\in \Sigma$ one can choose a coordinate system $(x,y,z)$ and a neighborhood $U\ni p$ such that
the intersection $W=U\cap \Sigma$ is formed by a graph $z=f(x,y)$ of a smooth function $f$ defined in an open domain of the $(x,y)$-plane.

\parbf{Examples.}
The simplest example of a surface is the $(x,y)$-plane 
\[\Pi=\set{(x,y,z)\in\RR^3}{z=0}.\]
The plane $\Pi$ is a surface since
it can be described as the graph of the function $f(x,y)=0$.

All other planes are surfaces as well since one can choose a coordinate system so that it becomes the $(x,y)$-plane.
We can also present a plane as a graph of a linear function 
$f(x,y)=a\cdot x+b\cdot y+c$ for some constants $a$, $b$ and $c$
(assuming the plane is not perpendicular to the $(x,y)$-plane).

A more interesting example is the unit sphere 
\[\SS^2=\set{(x,y,z)\in\RR^3}{x^2+y^2+z^2=1}.\]
This set is not the graph of any function,
but $\SS^2$ is locally a graph;
in fact it can be covered by 6 graphs:
\begin{align*}
z&=f_\pm(x,y)=\pm \sqrt{1-x^2-y^2},
\\
y&=g_\pm(x,z)=\pm \sqrt{1-x^2-z^2},
\\
x&=h_\pm(y,z)=\pm \sqrt{1-y^2-z^2};
\end{align*}
each function $f_\pm,g_\pm,h_\pm$ is defined in an open unit disc.
That is, $\SS^2$ is a smooth surface.

\parbf{More conventions.}
If the surface $\Sigma$ is formed by a closed set, then it is called \emph{complete}.
For example, paraboloids 
\[z=x^2+y^2,\quad\quad z=x^2-y^2\]
or sphere 
\[x^2+y^2+z^2=1\]
are complete surfaces, while the
open disc in a plane 
\[\set{(x,y,z)\in\RR^3}{x^2+y^2<1, z=0}\]
is a surface which is not complete.

If moreover $\Sigma$ is a compact set, then it is called \emph{closed surface} (the term \emph{closed set} is not directly relevant).

If a complete surface $\Sigma$ is noncompact, then it is called  \emph{open surface} (again the term \emph{open set} is not relevant).

For example, paraboloids 
are open surfaces, 
and sphere is closed.

A closed subset in a surface that is bounded by one or more smooth %???smooth???
curves is called \emph{surface with boundary}; in this case the collection of curves is called the \emph{boundary line} of the surface.
When we say \emph{surface} we usually mean a surface without boundary;
we may use the term \emph{surface with possibly nonempty boundary} if we need to talk about surfaces with and without boundary.

\section*{Local parametrizations}

Let $U$ be an open domain in $\RR^2$ and $f\:U\to \RR^3$ be a smooth map.
We say that $f$ is regular if its Jacobian has maximal rank;
that is, if the vectors $\tfrac{\partial f}{\partial u}$ and $\tfrac{\partial f}{\partial v}$ are linearly independent at any $(u,v)\in U$;
equivalently $\tfrac{\partial f}{\partial u}\times\tfrac{\partial f}{\partial v}\ne 0$, where $\times$ denotes the vector product.

\begin{thm}{Proposition}\label{prop:graph-chart}
If $f\:U\to \RR^3$ is a smooth regular embedding of an open connected set $U\subset \RR^2$, then it image $\Sigma=f(U)$ is a smooth surface.
\end{thm}

\parit{Proof.}
Let $f(u,v)=(f_1(u,v),f_2(u,v),f_3(u,v))$.
Since $f$ is regular the Jacobian
\[
\renewcommand\arraystretch{1.3}
\begin{pmatrix}
\tfrac{\partial f_1}{\partial u}&\tfrac{\partial f_1}{\partial v}\\
\tfrac{\partial f_2}{\partial u}&\tfrac{\partial f_2}{\partial v}\\
\tfrac{\partial f_3}{\partial u}&\tfrac{\partial f_3}{\partial v}
\end{pmatrix}
\]
has rank two.

Fix a point $p\in \Sigma$; by shifting the coordinate system we may assume that $p$ is the origin.
Permuting the coordinates $x,y,z$ if necessary, we may assume that 
the matrix 
\[
\renewcommand\arraystretch{1.3}
\begin{pmatrix}
\tfrac{\partial f_1}{\partial u}&\tfrac{\partial f_1}{\partial v}\\
\tfrac{\partial f_2}{\partial u}&\tfrac{\partial f_2}{\partial v}
\end{pmatrix}
\]
is invertible.
Let $\bar f\:U\to\RR^2$ be the projection of $f$ to the $(x,y)$-coordinate plane;
that is, $\bar f(u,v)=(f_1(u,v),f_2(u,v))$.
Note that the $2\times2$-matrix above is the Jacobian matrix of $\bar f$.

The inverse function theorem implies that there is a smooth regular function $h$ defined on an open set $W\ni 0$ in the $(x,y)$-plane
such that $h(0,0)=(0,0)$ and $\bar f\circ h$ is the identity map.

The graph $\Gamma$ described by $z=f_3\circ h(x,y)$ is a subset of $\Sigma$.
Indeed, if $(u,v)\z=h(x,y)$, then $x=f_1(u,v)$ and $y=f_2(u,v)$.
Therefore the identity $z=f_3\circ h(x,y)$ can be rewritten as $(x,y,z)=f(u,v)$.

Clearly $\Gamma$ is an open subset in $\Sigma$;
that is, $\Gamma$ a neighborhood of $p$ in $\Sigma$ that can be described as a graph of a smooth function $f_3\circ h\:W\to\RR$.
Since $p$ is arbitrary, we get that $\Sigma$ is a surface.
\qeds

If $f$ and $\Sigma$ as in the proposition, then we say that $f$ is a \emph{parametrization} of the surface $\Sigma$. 

Not all the smooth surfaces can be described by such a parametrization;
for example the sphere $\SS^2$ does not.
But any smooth surface $\Sigma$ admits a local parametrization; that is, any point $p\in\Sigma$ admits an open neighborhood $W\subset \Sigma$ with a smooth regular parametrization~$f$.
In this case any point in $W$ can be described by two parameters, usually denoted by $u$ and $v$, 
which are called \emph{local coordinates} at $p$.
The map $f$ is called a \emph{chart} of $\Sigma$.

If $W$ is a graph $z=h(x,y)$ then the map $f\:(u,v)\mapsto (u,v,h(u,v))$ is a chart.
Indeed, $f$ has an inverse $(u,v,h(u,v))\mapsto (u,v)$ which is continuous;
that is, $f$ is an embedding.
Further,
$\tfrac{\partial f}{\partial u}=(1,0,\tfrac{\partial h}{\partial u})$ and $\tfrac{\partial f}{\partial v}=(0,1,\tfrac{\partial h}{\partial v})$, whence $\tfrac{\partial f}{\partial u}$ and $\tfrac{\partial f}{\partial v}$ are linearly independent.

Note that from \ref{prop:graph-chart}, we obtain the following corollary.

\begin{thm}{Corollary}
A path connected set $\Sigma\subset \RR^3$ is a smooth regular surface if at any point $p\in \Sigma$ it has a local parametrization by a smooth regular map.
\end{thm}


\begin{thm}{Exercise}\label{ex:inversion}
Consider the following map 
\[f(u,v)=(\tfrac{2\cdot u}{1+u^2+v^2},\tfrac{2\cdot v}{1+u^2+v^2},\tfrac{2}{1+u^2+v^2}).\]
Show that $f$ is a chart of the unit sphere centered at $(0,0,1)$; describe the image of $f$.
\end{thm}

The map 
\[(u,v,1)\mapsto (\tfrac{2\cdot u}{1+u^2+v^2},\tfrac{2\cdot v}{1+u^2+v^2},\tfrac{2}{1+u^2+v^2})\]
is called \emph{stereographic projection}. 
Note that the point $(u,v,1)$ and its image $(\tfrac{2\cdot u}{1+u^2+v^2},\tfrac{2\cdot v}{1+u^2+v^2},\tfrac{2}{1+u^2+v^2})$ lie on one half-line starting at the origin.

Let $\gamma(t)=(x(t),y(t)$ be a plane curve.
Recall that the image of the map 
\[(t,\theta)\mapsto (x(t), y(t)\cdot\cos\theta,y(t)\cdot\sin\theta)\] 
is called \emph{surface of revolution} of the curve $\gamma$ around $x$-axis.

\begin{thm}{Exercise}\label{ex:revolution}
Assume $\gamma$ is a closed simple smooth regular plane curve that does not intersect $x$-axis.
Show that surface of revolution of $\gamma$ around $x$-axis is a smooth regular surface.
\end{thm}


\section*{Golbal parametrizations} 
A surface can be described by an embedding from a known surface to the space.
For example the ellipsoid
\[\Sigma_{a,b,c}=\set{(x,y,z)\in\RR^3}{\tfrac{x^2}{a^2}+\tfrac{y^2}{b^2}+\tfrac{z^2}{c^2}=1}\]
for some positive numbers $a,b,c$ can be defined as the image of the map $s\:\SS^2\to\RR^3$, defined as the restriction of the map $(x,y,z)\z\mapsto (a\cdot x, b\cdot y,c\cdot z)$ to the unit sphere $\SS^2$.

For a surface $\Sigma$, a map $s: \Sigma \to \RR ^3$ is called a 
\emph{smooth parametrized surface} if for for any chart $f\:U\to \Sigma$ 
the composition $s\circ f$ is smooth and regular;
that is all partial derivatives $\frac{\partial^{m+n}}{\partial u^m\partial v^n}(s\circ \tilde{f})$ exist and are continuous in the domain of definition and the following two vectors 
$\frac{\partial}{\partial u}(s\circ \tilde{f})$ $\frac{\partial}{\partial v}(s\circ \tilde{f})$ are linearly independent.

Evidently the parametric definition includes the embedded surfaces defined previously --- as the domain of parameters we can take the surface itself and the identity map as $s$,
but parametrized surfaces are more general, in particular they  might  have self-intersections.

If $\Sigma$ is a known surface for example a sphere or a plane, the paramtrized surface $s\:\Sigma\to\RR^3$ might be called by the same name.
For example any embedding $s\:\SS^2\to\RR^3$ might be called topological sphere
and if $s$ is smooth it might be called smooth sphere.
Similarly an embedding $s\:\RR^2\to\RR^3$ might be called topological plane
and if $s$ is smooth it might be called smooth plane.

\section*{Implicitly defined surfaces}

\begin{thm}{Proposition}
Let $f\:\RR^3\to \RR$ be a smooth function; 
that is, all its partial derivatives defined in its domain of definition.
Consider the set $\Sigma$ of solution of equation $f(x,y,z)=0$ in the plane.
Suppose that $\nabla f$ does not vanish at any point $p\in \Sigma$.
Then any path connected component of $\Sigma$ is a surface.
\end{thm}

\parit{Proof.}
Fix $p\in\Sigma$.
Since $\nabla_p f\ne 0$ we have 
\[\tfrac{\partial f}{\partial x}(p)\ne 0,\quad \tfrac{\partial f}{\partial y}(p)\ne 0,\quad \text{or}\quad\tfrac{\partial f}{\partial z}(p)\ne 0.\]
We may assume $\tfrac{\partial f}{\partial z}(p)\ne 0$;
otherwise permute the coordinates $x,y,z$.

The implicit function theorem (\ref{thm:imlicit}) implies that a neighborhood of $p$ in $\Sigma$ is the graph $z=h(x,y)$ of a smooth function $h$ defined on an open domain in $\RR^2$.
It remains to apply the definition of smooth surface (page \pageref{page:def-smooth-surface}).
\qeds

\begin{thm}{Exercise}\label{ex:hyperboloinds}
Describe the set of real numbers $a$
such that the equation
\begin{align*}
x^2+y^2-z^2&=a
\end{align*}
describes a smooth regular surface.
\end{thm}

\section*{Tangent plane}

Let $z=f(x,y)$ be a local graph realization of a surface. 
Assume $p=(x_p,y_p,z_p)$ lies on this graph, so $z_p=f(x_p,y_p)$.
The plane passing thru $p$ and spanned by the vectors $(1,0,(\tfrac{\partial}{\partial x}f)(x_p,y_p))$ and  $(0,1,(\tfrac{\partial}{\partial y}f)(x_p,y_p))$ is called the \emph{tangent plane} of $\Sigma$ at $p$.
It can be interpreted as the best approximation of the surface $\Sigma$ by a plane at~$p$.

The tangent plane to $\Sigma$ at $p$ is usually denoted by $\T_p$ or $\T_p\Sigma$.

It is straightforward to check that the tangent plane does not depend of the local presentation of $\Sigma$ by a graph.

\parbf{On local graph representations.}
The following proposition guarantees the existence of a local graph representation near a given point.

\begin{thm}{Proposition}\label{prop:perp}
Assume the tangent of a smooth surface $\Sigma$ at point $p$ is not perpendicular to the $(x,y)$-plane.
Then a neighborhood of $p$ in $\Sigma$ can be presented as a graph of smooth function $z=f(x,y)$ defined on an open set of the $(x,y)$-plane.
\end{thm}

A reader familiar with the inverse function theorem, can consider this proposition as an exercise.

\parbf{Tangent-normal coordinates.} 
Fix a point $p$ in a smooth surface $\Sigma$.
Consider a coordinate system $(x,y,z)$ with origin at $p$ such that the $(x,y)$-plane coincides with $\T_p$.

According to Proposition~\ref{prop:perp}, 
we can present $\Sigma$ locally around $p$ as a graph of a function $f$.
Note that $f$ satisfies the following additional properties:
\begin{align*}
f(0,0)&=0,
&
(\tfrac{\partial}{\partial x}f)(0,0)&=0,
&
(\tfrac{\partial}{\partial y}f)(0,0)&=0.
\end{align*}
The first equality holds since $p=(0,0,0)$ lies on the graph and the last two equalities mean that the tangent plane at $p$ is horizontal.

This gives an almost canonical coordinate system in a neighborhood of $p$;
it is unique up to a rotation of  the $(x,y)$-plane and switching the sign of the $z$-coordinate.

\section*{Curvatures}

\parbf{Hessian.}
Fix a point $p$ on a smooth surface $\Sigma$ and the associated tangent-normal coordinates. 

Consider the Hessian matrix 
\[M_p=\begin{pmatrix}
   (\tfrac{\partial^2}{\partial x^2}f)(0,0)
   &(\tfrac{\partial^2}{\partial x\partial y}f)(0,0)
   \\
   (\tfrac{\partial^2}{\partial y\partial x}f)(0,0)
   &(\tfrac{\partial^2}{\partial y^2}f)(0,0)
  \end{pmatrix}.
\]
This is a symmetric matrix, therefore by an appropriate rotation of the $(x,y)$-plane, we can make it diagonal;
that is, we can assume that $(\tfrac{\partial^2}{\partial x\partial y}f)(0,0)=0$.
Then the diagonal elements are called \emph{principle curvatures} of $\Sigma$ at $p$;
they are uniqueley defined up to sign;
They are denoted as $k_1(p)$ and $k_2(p)$.
The principle curvatures can be also defined as the eigenvalues of $M_p$.

The determinant of $M_p$ is $k_1(p)\cdot k_2(p)$;
it is called the \emph{Gauss curvature} of $\Sigma$ at $p$.
The trace of $M_p$ is $k_1(p)+ k_2(p)$;
it is called the \emph{mean curvature} of $\Sigma$ at $p$.

Form the discussion above, 
we get that the Gauss curvature depends only on $\Sigma$ and $p$,
and not on the choice of the coordinate system.
Up to sign, the same obsevation is true for the principle curvatures and the mean curvature. 


\begin{thm}{Exercise}\label{ex:projection}
Assume $\Sigma$ is a closed surface of with principle curvatures at most 1
and let $F$ be its orthogonal projection to a plane.
Show that no circle of curvature larger than 1 can support $F$ from the left. %???left???
\end{thm}

\begin{thm}{Exercise}
Show that any closed immersed surface has a point with positive Gauss curvature.
\end{thm}

\begin{thm}{Exercise}
Assume a closed surface $\Sigma$ bounds a convex body.
Show that $\Sigma$ is a sphere with nonnegative Gauss curvature. 
\end{thm}






